---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# derfinderPlot

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/derfinderPlot.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/derfinderPlot)
[![Codecov test coverage](https://codecov.io/gh/leekgroup/derfinderPlot/branch/master/graph/badge.svg)](https://codecov.io/gh/leekgroup/derfinderPlot?branch=master)
[![R build status](https://github.com/leekgroup/derfinderPlot/workflows/R-CMD-check/badge.svg)](https://github.com/leekgroup/derfinderPlot/actions)
[![R build status](https://github.com/leekgroup/derfinderPlot/workflows/test-coverage/badge.svg)](https://github.com/leekgroup/derfinderPlot/actions)
[![R build status](https://github.com/leekgroup/derfinderPlot/workflows/pkgdown/badge.svg)](https://github.com/leekgroup/derfinderPlot/actions)
<!-- badges: end -->

Addon package with plotting functions for
[derfinder](http://www.bioconductor.org/packages/derfinder) results.

## Documentation

For more information about `derfinderPlot` check the vignettes [through Bioconductor](http://bioconductor.org/packages/derfinderPlot) or at the [documentation website](http://leekgroup.github.io/derfinderPlot).

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `derfinderPlot` using from [Bioconductor](http://bioconductor.org/) the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("derfinderPlot")
```

## Citation

Below is the citation output from using `citation('derfinderPlot')` in R. Please 
run this yourself to check for any updates on how to cite __derfinderPlot__.

```{r 'citation', eval = requireNamespace('derfinderPlot')}
print(citation('derfinderPlot'), bibtex = TRUE)
```

Please note that the `derfinderPlot` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct
  
Please note that the derfinderPlot project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Development tools

* Testing on Bioc-devel is possible thanks to [GitHub actions through `r BiocStyle::CRANpkg('usethis')`](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/).
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh).
* The [documentation website](http://leekgroup.github.io/derfinderPlot) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
