---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# derfinderPlot

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![Bioc release status](http://www.bioconductor.org/shields/build/release/bioc/derfinderPlot.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/derfinderPlot)
[![Bioc devel status](http://www.bioconductor.org/shields/build/devel/bioc/derfinderPlot.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/derfinderPlot)
[![Bioc downloads rank](https://bioconductor.org/shields/downloads/release/derfinderPlot.svg)](http://bioconductor.org/packages/stats/bioc/derfinderPlot/)
[![Bioc support](https://bioconductor.org/shields/posts/derfinderPlot.svg)](https://support.bioconductor.org/tag/derfinderPlot)
[![Bioc history](https://bioconductor.org/shields/years-in-bioc/derfinderPlot.svg)](https://bioconductor.org/packages/release/bioc/html/derfinderPlot.html#since)
[![Bioc last commit](https://bioconductor.org/shields/lastcommit/devel/bioc/derfinderPlot.svg)](http://bioconductor.org/checkResults/devel/bioc-LATEST/derfinderPlot/)
[![Bioc dependencies](https://bioconductor.org/shields/dependencies/release/derfinderPlot.svg)](https://bioconductor.org/packages/release/bioc/html/derfinderPlot.html#since)
[![Codecov test coverage](https://codecov.io/gh/leekgroup/derfinderPlot/branch/devel/graph/badge.svg)](https://codecov.io/gh/leekgroup/derfinderPlot?branch=devel)
[![R build status](https://github.com/leekgroup/derfinderPlot/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/leekgroup/derfinderPlot/actions)
[![GitHub issues](https://img.shields.io/github/issues/leekgroup/derfinderPlot)](https://github.com/leekgroup/derfinderPlot/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/leekgroup/derfinderPlot)](https://github.com/leekgroup/derfinderPlot/pulls)
<!-- badges: end -->

Addon package with plotting functions for
[derfinder](http://www.bioconductor.org/packages/derfinder) results.

## Documentation

For more information about `derfinderPlot` check the vignettes [through Bioconductor](http://bioconductor.org/packages/derfinderPlot) or at the [documentation website](http://leekgroup.github.io/derfinderPlot).

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `derfinderPlot` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("derfinderPlot")
```

## Citation

Below is the citation output from using `citation('derfinderPlot')` in R. Please 
run this yourself to check for any updates on how to cite __derfinderPlot__.

```{r 'citation', eval = requireNamespace('derfinderPlot')}
print(citation("derfinderPlot"), bibtex = TRUE)
```

Please note that the `derfinderPlot` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct
  
Please note that the derfinderPlot project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, `r BiocStyle::Githubpkg('r-hub/sysreqs')` and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://leekgroup.github.io/derfinderPlot) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.
